The Journey from Wild to Text Book Data: A Case Study of the National Longitudinal Survey of Youth

### 2021-08-18 Report/report_alt.Rmd
Modify the EDA.

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE,
                      #cache = TRUE, 
                      #cache.path = "cache/",
                      fig.path = "figures/")
library(tidyverse)
library(brolgar)
library(patchwork)
library(kableExtra)
library(MASS)
library(janitor)
library(DiagrammeR)
library(rsvg)
library(webshot)
library(mgcv)
library(modelr)
library(tsibble)
library(lme4)
library(modelr)
```

@@

wages_slope <- key_slope(wages_hs2020_tsibble, 
            mean_hourly_wage ~ year) %>%
  left_join(demographic_nlsy79, by = "id") %>%
  mutate(hgc = factor(hgc, levels = c("1ST GRADE", "2ND GRADE", "3RD GRADE", 
                                      "4TH GRADE", "5TH GRADE", "6TH GRADE", 
                                      "7TH GRADE", "8TH GRADE", "9TH GRADE",
                                      "10TH GRADE", "11TH GRADE", "12TH GRADE", 
                                      "UNGRADED"))) %>%
  filter(hgc != "UNGRADED")

ggplot(wages_slope, aes(y = hgc,
                      x = .slope_year)) +
  geom_boxplot()
```

@@

# transform the data type and year for modeling
wages_model <- wages_hs2020_tsibble %>%
  mutate(id_fct = as.factor(id),
         year0 = year - 1979,
         ln_wages = log(mean_hourly_wage))

# build GAM model
wages_fit <- lmer(
  ln_wages ~ year0 + hgc_i + gender + (1|id),
  data = wages_model
)

wages_fit_aug<- wages_model %>%
  add_predictions(wages_fit, var = "pred") %>%
  add_residuals(wages_fit, var = "resid")

slope_only <- wages_slope %>%
  dplyr::select(id, .slope_year) %>%
  arrange(desc(.slope_year)) %>%
  slice_head(n = 12)

wages_fit_aug2 <- wages_fit_aug %>%
  filter(id %in% slope_only$id)

>>>>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE,
                      #cache = TRUE, 
                      #cache.path = "cache/",
                      fig.path = "figures/")
library(tidyverse)
library(brolgar)
library(patchwork)
library(kableExtra)
library(MASS)
library(janitor)
library(DiagrammeR)
library(rsvg)
library(webshot)
library(mgcv)
library(modelr)
library(tsibble)
library(lme4)
library(modelr)
library(forcats)
```

@@

# transform the data for modeling purpose
wages_model_data <- wages_hs2020_tsibble %>%
  mutate(id_fct = as.factor(id),
         year0 = year - 1979,
         ln_wages = log(mean_hourly_wage)) %>%
  mutate(hgc_regroup = fct_collapse(hgc, `8TH grade or less` = c("UNGRADED","1ST GRADE", "3RD GRADE",
                                                                 "4TH GRADE", "5TH GRADE",
                                                                 "6TH GRADE", "7TH GRADE",
                                                                 "8TH GRADE"),
                                    `9TH to 11TH grade` = c("9TH GRADE",
                                                            "10TH GRADE",
                                                            "11TH GRADE"),
                                    `12TH grade` = "12TH GRADE"))

# extract the slope for every individual 
wages_slope <- key_slope(wages_model_data, 
            ln_wages ~ year0) %>%
  left_join(demographic_nlsy79, by = "id") %>%
  mutate(hgc_regroup = fct_collapse(hgc, `8TH grade or less` = c("UNGRADED","1ST GRADE", "3RD GRADE",
                                                                 "4TH GRADE", "5TH GRADE",
                                                                 "6TH GRADE", "7TH GRADE",
                                                                 "8TH GRADE"),
                                    `9TH to 11TH grade` = c("9TH GRADE",
                                                            "10TH GRADE",
                                                            "11TH GRADE"),
                                    `12TH grade` = "12TH GRADE"))

ggplot(wages_slope) +
  geom_boxplot(aes(x = `.slope_year0`, y = hgc_regroup))

```
```{r}
ggplot(wages_slope, aes(x = .slope_year0)) +
  geom_histogram()
```

```{r}
slope_bottom_5<- wages_slope %>%
  arrange(.slope_year0) %>%
  slice_head(n = 5)

slope_top_5<- wages_slope %>%
  arrange(desc(.slope_year0)) %>%
  slice_head(n = 5)

keep_contrast <- rbind(slope_bottom_5, slope_top_5) %>%
  dplyr::select(.slope_year0, id)

contrast_slope <- left_join(keep_contrast, wages_model_data, by = "id") %>%
  arrange(.slope_year0) %>%
  mutate(id = factor(id, levels = c("7918", "11695", "9176", "10119", "6464", "1271", 
                                    "12371", "9039", "10217", "7782")))

ggplot(contrast_slope, aes(x = year, y = ln_wages, color = gender)) +
  geom_line() +
  geom_point() +
  facet_wrap(~id)

```{r}
ggplot(contrast_slope, aes(x = year, y = ln_wages, color = hgc_regroup)) +
  geom_line() +
  geom_point() +
  facet_wrap(~id)
```

```{r}
quartile <- wages_slope %>%
  mutate(quartile = as.factor(ntile(.slope_year0, 4)))

ggplot(quartile) +
  geom_bar(aes(x = quartile, fill = gender, position = "fill"))
```

```{r}
ggplot(quartile) +
  geom_bar(aes(x = quartile, fill = hgc_regroup, position = "fill"))
```

@@

# build linear-mixed model
wages_fit <- lmer(
  ln_wages ~ year0*hgc_regroup*gender + (1|id),
  data = wages_model_data
)

wages_fit_aug<- wages_model_data %>%
  add_predictions(wages_fit, var = "pred") %>%
  add_residuals(wages_fit, var = "resid")

slope_bottom_12 <- wages_slope %>%
  dplyr::select(id, .slope_year0) %>%
  arrange(.slope_year0) %>%
  slice_head(n = 12)

wages_fit_aug2 <- wages_fit_aug %>%
  filter(id %in% slope_bottom_12$id)

### 2021-08-18 Report/report_alt.tex
Calculate gender wage gap and fit linear mixed model.
Modify the EDA.

In the second example, we examined the pattern of mean hourly wage corresponding to different groups of gender and race. We took 240 samples of females and males equally and spread them into twelve facets, as shown in Figure \ref{fig:plot-gender} using \texttt{facet\_strata} function from \texttt{brolgar} (Tierney, Cook, and Prvan 2020). We learn that the mean hourly wage of females and males fluctuated over the years. For example, in facets 4, 5, and 10, males tend to earn more wages than females. This finding is also shown in Figure \ref{fig:se-gender} where we plotted the median wage of males and females against the year. We learn that males had a higher mean of mean hourly wage throughout the survey than females. Unfortunately, the gap between them tends to be wider from time to time.

@@

\begin{figure}
\centering
\includegraphics{figures/se-gender-1.pdf}
\caption{\label{fig:se-gender}The yearly mean of mean hourly wage by gender along with its standard error. Males generally earn more wages than females. The error interval tend to be wider over the years.}
\end{figure}

For the final example, we showed modeling the longitudinal data using the Generalized Additive Model (GAM) with \texttt{mgcv} package (Wood 2003) for each ID. We fitted a more flexible model to fit into the data since the relationship between mean hourly wage and year is not linear due to the spikes. Since this approach is computationally expensive, we only showed the model in a small fraction of the data. We sampled 1 percent of the total respondents using \texttt{brolgar\textquotesingle{}s} \texttt{sample\_frac\_keys} function. Further, we plotted some IDs overlayed by the fitted GAM model in Figure \ref{fig:plot-gam}. We learn that the model flexible enough to deal with the fluctuation in the data. Moreover, the data with small spikes tend to be linearly fitted.

@@

@@

>>>>

In the second example, we examined the pattern of mean hourly wage corresponding to different groups of gender and race. We took 240 samples of females and males equally and spread them into twelve facets, as shown in Figure \ref{fig:plot-gender} using \texttt{facet\_strata} function from \texttt{brolgar} (Tierney, Cook, and Prvan 2020). We learn that the mean hourly wage of females and males fluctuated over the years. For example, in facets 4, 5, and 10, males tend to earn more wages than females.

@@

To further investigate this, we calculated the median wage of females relative to males, which is called gender wage gap. According to OECD (n.d.), gender pay gap is the difference between median wage of males and females relative to median wage of males. We find that, as shown in \ref{fig:se-gender}, over the years, females tend to earn fewer than males. For example, in 2018 females earn about 78 cents for every \$1 earned by males. The widest gender gaps were observed in 1990s.

\begin{figure}
\centering
\includegraphics{figures/se-gender-1.pdf}
\caption{\label{fig:se-gender}Gender wage gap or the median wages of females relatives to median wages of males. Females tend to earn fewer than males. For example, in 2018 females earn about 78 cents for every \$1 earned by males.}
\end{figure}

For the final example, we showed modeling the longitudinal data using linear mixed model with \texttt{lme4} package (Bates et al. 2015).

We fitted a more flexible model to fit into the data since the relationship between mean hourly wage and year is not linear due to the spikes. Since this approach is computationally expensive, we only showed the model in a small fraction of the data. We sampled 1 percent of the total respondents using \texttt{brolgar\textquotesingle{}s} \texttt{sample\_frac\_keys} function. Further, we plotted some IDs overlayed by the fitted GAM model in Figure \ref{fig:plot-gam}. We learn that the model is flexible enough to deal with the fluctuation in the data. Moreover, the data with small spikes tend to be linearly fitted.

@@

\leavevmode\hypertarget{ref-lme4}{}%
Bates, Douglas, Martin Mächler, Ben Bolker, and Steve Walker. 2015. ``Fitting Linear Mixed-Effects Models Using lme4.'' \emph{Journal of Statistical Software} 67 (1): 1--48. \url{https://doi.org/10.18637/jss.v067.i01}.

@@

\leavevmode\hypertarget{ref-OECD}{}%
OECD. n.d. ``Gender Wage Gap.'' https://data.oecd.org/earnwage/gender-wage-gap.htm.
